Staged Abstract Interpreters

### 2019-09-16 paper/20_prelim.tex
polish

\paragraph{Binding-time Abstraction}
The binding-time type is declared as a higher-kinded type @R[_]@ \cite{Ofenbeck:2017:SGP:3136040.3136060}.
Now we can use @R@ to annotate other data types in the interpreter.
Later, the binding-time information is also injected into the monadic type
@MonadOps@.  If we instantiate @R@ as the identity type (i.e., @type R[T] = T@), 
then the generic interpreter is a standard definitional interpreter that
will execute the program.  In Section \ref{stagedinterp}, we instantiate @R@
using LMS's built-in next-stage type annotation @Rep@, which makes the
interpreter act as a compiler.

@@

We also declare several operations to manipulate environments and
stores. These methods return monadic values of type @AnsM[_]@, which
may be parameterized over the environment or store type, or merely a @Unit@
value for performing effects. For example, @local_env@ installs a new environment
@ρ@ when evaluating the monadic value @ans@; @set_store@ takes a pair of
addresses and (potentially staged) values, and updates the store accordingly.
\begin{lstlisting}
  type MonadOps[R[_], M[_], A] = {           // Environment operations
    def map[B](f: R[A] => R[B]): M[B]        def ask_env: AnsM[Env]
    def flatMap[B](f: R[A] => M[B]): M[B]    def local_env(ans: Ans)(ρ: R[Env]): Ans
  }                                          // Store operations
                                             def get_store: AnsM[Store]
  type AnsM[T] <: MonadOps[R, AnsM, T]       def put_store(σ: R[Store]): AnsM[Unit]
  type Ans = AnsM[Value]                     def set_store(av: (R[Addr], R[Value])): AnsM[Unit]
\end{lstlisting}

\paragraph{Primitive Operations} Next, we define a few primitive operations.
First, we declare two versions of @alloc@. The first takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is
consequently wrapped by @R[_]@. The other @alloc@ is simply the monadic version
of the addresses, and can therefore be used in monadic computations.
An auxiliary method @get@ retrieves the value of an identifier @x@ through the
environment and store.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
\end{lstlisting}

Other primitive operations in the interpreter handle the language constructs.  The
methods @num@ and @close@ deal with primitive values, which lift literal terms
(e.g., lambdas) to our value representation (e.g., closures).
Conditionals and arithmetic are handled by @br0@ and @arith@, respectively. The
method @ap_clo@ is used for applying functions, by taking a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are all
annotated by @R[_]@, as they are potentially next-stage values when the
interpreter acts as a compiler.
\begin{lstlisting}
  def num(i: Int): Ans
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(fun: R[Value], arg: R[Value]): Ans
\end{lstlisting}

>>>>

\paragraph{Binding-time Abstraction}
The binding-time type is declared as a higher-kinded type @R[_]@ \cite{Ofenbeck:2017:SGP:3136040.3136060}.
Now we can use @R@ to annotate other data types in the interpreter.
Later, the binding-time information is also injected into the monadic type
@MonadOps@.  If we instantiate @R@ as the identity type (i.e., @type R[T] = T@), 
then the generic interpreter is a standard definitional interpreter that
will execute the program.  In Section \ref{stagedinterp}, we instantiate @R@
using LMS's built-in next-stage type annotation @Rep@, which makes the
interpreter act as a compiler.
\begin{lstlisting}
  type MonadOps[R[_], M[_], A] = {           // Environment operations
    def map[B](f: R[A] => R[B]): M[B]        def ask_env: AnsM[Env]
    def flatMap[B](f: R[A] => M[B]): M[B]    def local_env(ans: Ans)(ρ: R[Env]): Ans
  }                                          // Store operations
                                             def get_store: AnsM[Store]
  type AnsM[T] <: MonadOps[R, AnsM, T]       def put_store(σ: R[Store]): AnsM[Unit]
  type Ans = AnsM[Value]                     def set_store(av: (R[Addr], R[Value])): AnsM[Unit]
\end{lstlisting}

@@

We also declare several operations to manipulate environments and
stores. These methods return monadic values of type @AnsM[_]@, which
may be parameterized over the environment or store type, or merely a @Unit@
value for performing effects. For example, @local_env@ installs a new environment
@ρ@ when evaluating the monadic value @ans@; @set_store@ takes a pair of
addresses and (potentially staged) values, and updates the store accordingly.

\paragraph{Primitive Operations} Next, we declare a few primitive operations.
First, we declare two versions of @alloc@. The first takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is
consequently wrapped by @R[_]@. The other @alloc@ is simply the monadic version
of the previous one, and can therefore be used in monadic computations.
An auxiliary method @get@ retrieves the value of an identifier @x@ through the
environment and store.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
\end{lstlisting}

Other primitive operations in the interpreter handle the language constructs.  The
methods @num@ and @close@ deal with primitive values. For example, @close@ lifts literal terms
(e.g., lambdas) to our value representation (e.g., closures).
Conditionals and arithmetic are handled by @br0@ and @arith@, respectively. The
method @ap_clo@ is used for applying functions, by taking a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are all
annotated by @R[_]@, as they are potentially next-stage values when the
interpreter acts as a compiler.
\begin{lstlisting}
  def num(i: Int): Ans
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(fun: R[Value], arg: R[Value]): Ans
\end{lstlisting}
